HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-08-06 scnn.tex
update in result section

To improve the stability of the spherical CNN, we slowly increase the amount of noise added to the training data over the iterations.
Furthermore in order to avoid overfitting, the noise is randomly produced during the training process.
We trained our model between $50$ and $290$ epochs depending on the complexity and the size of the problem.
The $20\%$ of the training set that we call \emph{validation set} was left out of the training process and simply used to assess the global performance of the network.
Training took less than a day for each model using a single NVIDIA Tesla K20.

We compare the performance of the spherical CNN against two simple benchmark algorithms.
The two baselines are based on building features with
(a) the Power Spectral Densities (PSD) and
(b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified using a Support Vector Machine (SVM) with a linear kernel.
Other kernels did not provide better results, while having significantly worse scaling properties.
All classifiers are trained with noisy samples.
For fairness, we augmented the dataset of the baseline algorithm in a similar way as for the network; we created samples by adding new random realizations of the noise.
We stop adding new samples in the classifier once the validation error converges.
This process is detailed in \ref{sec:dataset_augmentation}.
We tune the regularization parameter by cross-validating with the validation set.

@@

\todo{We use the ADAM optimizer, \cite{kingma2014adam}, with an initial learning rate of $2e-4$ that is exponentially decreased by $0.98$ a $100$ times to end at $2.7e-5$.}

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\linewidth]{result_order1}
\hfill
\includegraphics[width=0.32\linewidth]{result_order2}
\hfill
\includegraphics[width=0.32\linewidth]{result_order4}
\caption{
Classification errors for the 3 different areas of the sphere.
Green lines show the performance of the CNN, blue of the PDF+SVM, and red of the histogram+SVM.
Green dashed line corresponds to the performance of CNN without the statistical layer.
The noise level was increased from zero to $2\times$ the standard deviation of the mass map.
}
\label{fig:results}
\end{figure*}

@@

We observe that the PSD is unable to perform the desired task.
The main reason is that the two different models were designed to have very similar PSDs (see \figref{psd_sigma3}).
As a result, one needs to use other statistics to solve the problem.
In this case, the histogram features allow for a significant improvement over the PSD ones.
Unfortunately, while the histogram features contain information about the distribution of pixels, they do not include any spatial information.
In contrast, both the PSD features and the HealpixNet can exploit the spatial information.
Note that without this graph structure, a fully connected neural network approach would have to be used.
Problematically, fully connected neural networks generally scales squarely with their number of inputs.
\TK{I don't understand this comment.} \nati{Maybe we should leave it out then. It is not that important.}

We tried several different spherical neural network architectures and observed the following results.
First, Chebyshev polynomial seems more stable than monomials ($\theta_k$ in \eqnref{graph_convolution_polynomial}), in the sense that the training loss had a more steadily decrease.
%\TK{what are monomials? stable in what sense?}
Nevertheless once the network is trained, we could not observe a significant accuracy difference.
Second, using a statistical layer improved the result unlimited for our problem.
We tried $4$ different layers `mean', `variance', `mean + variance' and `histogram'.
The `mean', `histogram' and `mean + variance' were providing the same level of accuracy, with pure `variance' being slightly worse.
Third, the use of the dropout\TK{was dropout mentioned before?} \nati{Probably not. But to keep it simple, we may just add a footnote or a reference.} layer did not help in the presence of noise.
This is probably due to the fact that the dataset was augmented indefinitely through the addition of noise.
Fourth, especially for the more difficult cases, orders 2 or 4 and large noise levels, a large architecture with many filters and many training iterations were necessary to achieve high accuracy levels.

@@

In this contribution we presented HealpixNet: a spherical convolutional neural network for the HEALPix sampling.
The core of our method is the use of a graph to represent the pixel grid on the sphere.
This allows us to leverage the advantages of the graph convolution; it is both efficient, with a complexity of $\mathcal{O}(N_{pix})$, and flexible, which allows to work only on a part of the sphere.
Spherical properties of the domain are still captured well.
The method presented here restricts the shape of the filters learned by the network to be radial.
While this may be a less general design compared to the one used by conventional CNNs on flat images, we found it to still work efficiently for our test cases.
The layered structure of the network can partially compensate for that.
Further generalizations to non-radial filters may be possible [cite?]\TK{update here}, but we leave it to future work.

>>>>

% Nati: Not done anymore
% To improve the stability of the spherical CNN, we slowly increase the amount of noise added to the training data over the iterations.
Furthermore in order to avoid overfitting, the noise is randomly produced during the training process.
The $20\%$ of the training set that we call \emph{validation set} was left out of the training process and simply used to assess the global performance of the network.

We compare the performance of the spherical CNN against two simple benchmark algorithms.
The two baselines are based on building features with
(a) the Power Spectral Densities (PSD) and
(b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified using a Support Vector Machine (SVM) with a linear kernel.
Other kernels did not provide better results, while having significantly worse scaling properties.
All classifiers are trained with noisy samples.
For fairness, we augmented the dataset of the baseline algorithm in a similar way as for the network; we created samples by adding new random realizations of the noise.
We stop adding new samples in the classifier once the validation error converges.
This process is detailed in \ref{sec:dataset_augmentation}.
We tune the SVM regularization parameter by cross-validating with the validation set.

@@

\todo{We use the ADAM optimizer, \cite{kingma2014adam}, with an initial learning rate of $2e-4$ that is exponentially decreased by $0.98$ a $100$ times to end at $2.7e-5$.
All models where trained $80$ epochs with a batch size of $16\times o^2$. With this choice, all models recieve an identical amount of data within a batch and were trained with $1920$ gradient steps.}
Training took less \nati{queue bottleneck to be fixed} than $3$ hours for each model using a single NVIDIA Tesla K20.

@@

\begin{figure*}
\centering
\includegraphics[width=0.32\linewidth]{result_order1}
\hfill
\includegraphics[width=0.32\linewidth]{result_order2}
\hfill
\includegraphics[width=0.32\linewidth]{result_order4}
\caption{
Classification errors for the 3 different areas of the sphere.
Green lines show the performance of the CNN, blue of the PDF+SVM, and red of the histogram+SVM.
% Green dashed line corresponds to the performance of CNN without the statistical layer.
The noise level was increased from zero to $2\times$ the standard deviation of the mass map.
}
\label{fig:results}
\end{figure*}

@@

We observe that the PSD is unable to perform the desired task.
The main reason is that the two different models were designed to have very similar PSDs (see \figref{psd_sigma3}).
As a result, one needs to use other statistics to solve the problem.
In this case, the histogram features allow for a significant improvement over the PSD ones.
Unfortunately, while the histogram features contain information about the distribution of pixels, they do not include any spatial information.
In contrast, both the PSD features and the HealpixNet can exploit the spatial information.
% Note that without this graph structure, a fully connected neural network approach would have to be used.
% Problematically, fully connected neural networks generally scales squarely with their number of inputs.
% \TK{I don't understand this comment.} \nati{Maybe we should leave it out then. It is not that important.}

We tried several different spherical neural network architectures and observed the following results.
First, Chebyshev polynomial seems more stable than monomials ($\theta_k$ in \eqnref{graph_convolution_polynomial}), in the sense that the training loss had a more steadily decrease.
Nevertheless once the network is trained, we could not observe a significant accuracy difference.
Second, using a statistical layer slightly improved the accuracy in our problem.
We tried $4$ different layers `mean', `variance', `mean + variance' and `histogram'.
The `mean', `histogram' and `mean + variance' were providing the same level of accuracy, with pure `variance' being slightly worse. Hence we chose to simply use the `mean' statistic.
Third, adding fully connected layers after the statistical layer does not improve the performance, but lead to models requiring more training steps and regularization.
\todo{Fourth, using $\ell_2$ regularization does not help neither with the performance nor the training of the models, but slightly reduce the performance variance between different runs due to the random initialization.
Eventually, recommend to ignitialize the Chebyshev coeffecient with a centered Gaussian distribution with standart deviation $2/(F * (K + 0.5))$, where $K$ is the Chebyshef order and $F$ the number of input channels. The
}

@@

In this contribution we presented HealpixNet: a spherical convolutional neural network for the HEALPix sampling.
The core of our method is the use of a graph to represent the pixel grid on the sphere.
This allows us to leverage the advantages of the graph convolution; it is both efficient, with a complexity of $\mathcal{O}(N_{pix})$, and flexible, which allows to work only on a part of the sphere.
Spherical properties of the domain are still captured well.
The method presented here restricts the shape of the filters learned by the network to be radial.
While this may be a less general design compared to the one used by conventional CNNs on flat images, we found it to still work efficiently for our test cases.
The layered structure of the network can partially compensate for that.
% Further generalizations to non-radial filters may be possible [cite?]\TK{update here}, but we leave it to future work.
